---
title: "SBA 7(a) Loan Data — Exploratory Analysis"
author: "Colin McBane"
date: today
format: html
execute:
  echo: false
  warning: false
  message: false
---

## Overview

This report analyzes raw SBA 7(a) federal loan data cleaned in Phase 1. The goal is to understand the distribution of loan characteristics, identify default patterns, and inform feature selection for Phase 4 modeling.

```{r setup}
#| label: setup
#| message: false
#| warning: false

# Load libraries
library(tidyverse)
library(DBI)
library(RSQLite)
library(ggplot2)
library(corrplot)
library(scales)

# Connect to SQLite database built in Phase 1
conn <- dbConnect(
  RSQLite::SQLite(),
  "data/processed/sba_clean.db"
)

# Load the clean loans view into R
loans <- dbGetQuery(conn, "SELECT * FROM loans_clean")

# Quick sanity check
cat("Rows loaded:", nrow(loans), "\n")
cat("Columns:", ncol(loans), "\n")
cat("Default rate:", round(mean(loans$is_default) * 100, 2), "%\n")
```

## Loan Amount Distribution

```{r loan-distribution}
#| label: loan-distribution
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 6

# Loan amount distribution by default status
ggplot(loans, aes(x = loan_amount, fill = factor(is_default))) +
  geom_histogram(bins = 50, alpha = 0.7, position = "identity") +
  scale_x_continuous(
    labels = dollar_format(scale = 1e-6, suffix = "M"),
    limits = c(0, 2000000)
  ) +
  scale_fill_manual(
    values = c("0" = "#2196F3", "1" = "#F44336"),
    labels = c("0" = "Paid in Full", "1" = "Charged Off")
  ) +
  labs(
    title = "SBA 7(a) Loan Amount Distribution by Default Status",
    subtitle = "FY2010–Present | 489,935 loans",
    x = "Loan Amount",
    y = "Number of Loans",
    fill = "Outcome"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "top"
  )
```

**Key Finding:** SBA 7(a) loans are heavily concentrated below \$250,000 with a strong right skew. Charged-off loans appear proportionally more frequent in smaller loan buckets, which may suggest higher relative default risk at the lower end. However this is correlation not causation — larger loans undergo more rigorous underwriting scrutiny before approval, meaning the lower default rate on large loans likely reflects selection bias in the approval process rather than loan size itself being protective. This distinction will be important when interpreting model coefficients in Phase 4.

## Default Rate by Industry

```{r default-by-industry}
#| label: default-by-industry
#| fig-width: 12
#| fig-height: 7

# Pull industry summary from SQLite
industry <- dbGetQuery(conn, "
  SELECT naics_sector, industry_description, 
         total_loans, default_rate_pct
  FROM summary_by_industry
  WHERE total_loans > 500
  ORDER BY default_rate_pct DESC
  LIMIT 15
")

# Clean up industry descriptions for chart labels
# Clean up industry descriptions for chart labels
industry <- industry %>%
  group_by(industry_description) %>%
  slice_max(default_rate_pct, n = 1) %>%
  ungroup() %>%
  mutate(industry_description = str_wrap(industry_description, width = 40)) %>%
  distinct(industry_description, .keep_all = TRUE) %>%
  arrange(default_rate_pct) %>%
  mutate(industry_description = factor(industry_description,
                                        levels = unique(industry_description)))

# Plot
ggplot(industry, aes(x = industry_description, 
                      y = default_rate_pct,
                      fill = default_rate_pct)) +
  geom_col() +
  geom_text(aes(label = paste0(default_rate_pct, "%")),
            hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_fill_gradient(low = "#FFF9C4", high = "#B71C1C") +
  scale_y_continuous(limits = c(0, max(industry$default_rate_pct) * 1.15)) +
  labs(
    title = "Top 15 Industries by SBA Loan Default Rate",
    subtitle = "Industries with 500+ loans | Ranked by default rate",
    x = NULL,
    y = "Default Rate (%)",
    fill = "Default Rate"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "none"
  )
```

**Key Finding:** Limousine and personal transportation services show the highest default rates at nearly 19%, followed by residential remodeling and electronic retail. These industries share common risk characteristics — project-based or discretionary revenue, high sensitivity to economic cycles, and thin operating margins. This finding will inform industry-level risk features in Phase 4 modeling.

## Default Rate by State

```{r default-by-state}
#| label: default-by-state
#| fig-width: 12
#| fig-height: 7

# Pull state summary from SQLite
states <- dbGetQuery(conn, "
  SELECT borr_state, total_loans, default_rate_pct
  FROM summary_by_state
  WHERE total_loans > 200
  AND borr_state NOT IN ('', 'NULL')
  ORDER BY default_rate_pct DESC
  LIMIT 20
")

# Sort for chart
states <- states %>%
  arrange(default_rate_pct) %>%
  mutate(borr_state = factor(borr_state, levels = borr_state))

# Plot
ggplot(states, aes(x = borr_state,
                    y = default_rate_pct,
                    fill = default_rate_pct)) +
  geom_col() +
  geom_text(aes(label = paste0(default_rate_pct, "%")),
            hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_fill_gradient(low = "#E3F2FD", high = "#1565C0") +
  scale_y_continuous(limits = c(0, max(states$default_rate_pct) * 1.15)) +
  labs(
    title = "Top 20 States by SBA Loan Default Rate",
    subtitle = "States with 200+ loans | Ranked by default rate",
    x = NULL,
    y = "Default Rate (%)",
    fill = "Default Rate"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "none"
  )
```

**Key Finding:** Florida leads all states with an 11.74% default rate, followed by DC and New Jersey at roughly 10.6% and 10.6% respectively. Florida's elevated default rate is consistent with its exposure to natural disaster risk — hurricanes and flooding can devastate small businesses overnight, disrupting cash flow and making loan repayment impossible regardless of the borrower's original creditworthiness. This is a systemic geographic risk that individual underwriters cannot control at the loan level.

DC, New Jersey, and New York all appear in the top five, which aligns with high cost-of-living environments where small business operating costs — rent, labor, utilities — compress margins and reduce the buffer against economic shocks. A restaurant in Manhattan faces fundamentally different fixed cost pressure than the same restaurant in rural Tennessee.

Texas and California, despite their large loan volumes, show moderate default rates around 9-10%, suggesting that economic diversification at the state level provides some natural hedge against industry-specific downturns.

This geographic signal will be incorporated as a state-level risk feature in Phase 4 modeling. It also directly connects to Phase 5 fairness analysis — states with structurally higher default rates may create disparate impact concerns if geography is used as a direct model input.

## Default Rate Over Time

```{r default-by-year}
#| label: default-by-year
#| fig-width: 11
#| fig-height: 6

# Pull yearly summary from SQLite
yearly <- dbGetQuery(conn, "
  SELECT approval_year, total_loans, 
         default_rate_pct, avg_loan_amount
  FROM summary_by_year
  WHERE approval_year IS NOT NULL
  ORDER BY approval_year
")

# Dual axis plot — default rate and loan volume over time
ggplot(yearly, aes(x = approval_year)) +
  geom_col(aes(y = total_loans / 10),
           fill = "#E3F2FD", color = "#90CAF9", alpha = 0.7) +
  geom_line(aes(y = default_rate_pct * 100),
            color = "#B71C1C", linewidth = 1.2) +
  geom_point(aes(y = default_rate_pct * 100),
             color = "#B71C1C", size = 3) +
  scale_x_continuous(breaks = seq(min(yearly$approval_year),
                                   max(yearly$approval_year), 1)) +
  scale_y_continuous(
    name = "Loan Volume (scaled)",
    sec.axis = sec_axis(~ . / 100,
                        name = "Default Rate (%)",
                        labels = percent_format())
  ) +
  labs(
    title = "SBA 7(a) Loan Volume and Default Rate by Approval Year",
    subtitle = "Blue bars = loan volume | Red line = default rate",
    x = "Approval Year"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

**Key Finding:** This chart reveals three distinct macroeconomic periods that shaped SBA 7(a) lending patterns between 2010 and 2026.

**2010–2019 — Post-Crisis Recovery:** Loan volume grew steadily through the decade as the economy recovered from the 2008 financial crisis. Default rates remained relatively stable in the 1,000–1,500 basis point range, reflecting a healthy but cautious lending environment.

**2020–2021 — COVID Displacement:** Loan volume dropped sharply in 2020 as the SBA redirected resources toward emergency programs — the Paycheck Protection Program (PPP) and Economic Injury Disaster Loans (EIDL) — which provided direct grants and forgivable loans to small businesses. These programs temporarily absorbed demand that would otherwise have flowed through the 7(a) program. Critically, default rates did not spike immediately during the pandemic because government relief programs artificially supported borrower cash flows, masking underlying credit stress.

**2022–2023 — The Macroeconomic Double Whammy:** Default rates peaked in this period as the Federal Reserve aggressively raised interest rates to combat post-pandemic inflation — the fastest rate hiking cycle since the 1980s. Borrowers who took on variable-rate SBA loans during the low-rate environment of 2020–2021 suddenly faced materially higher debt service costs. Simultaneously, inflation compressed operating margins as input costs rose faster than businesses could pass through price increases to customers. The combination of higher rates and margin compression created a perfect storm for small business defaults — explaining why the default spike lagged the pandemic by two years rather than occurring during it.

**2024–2026 — Right Censoring:** Default rates appear to fall sharply in recent years, but this is a statistical artifact rather than a genuine improvement in credit quality. Loans approved in 2023–2025 have not had sufficient time to season and default. In credit risk modeling this is called right censoring — recent observations are incomplete because the outcome has not yet been observed. Phase 4 modeling will account for this by weighting or excluding recent vintages from the training set to avoid biasing the model toward artificially low default rates in recent years.

## Interest Rate vs Default Status

```{r interest-rate-default}
#| label: interest-rate-default
#| fig-width: 10
#| fig-height: 6

# Box plot of interest rate by default status
ggplot(loans, aes(x = factor(is_default),
                   y = interest_rate,
                   fill = factor(is_default))) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.2) +
  geom_violin(alpha = 0.3) +
  scale_fill_manual(
    values = c("0" = "#2196F3", "1" = "#F44336"),
    labels = c("0" = "Paid in Full", "1" = "Charged Off")
  ) +
  scale_x_discrete(labels = c("0" = "Paid in Full", "1" = "Charged Off")) +
  scale_y_continuous(limits = c(0, 20)) +
  labs(
    title = "Interest Rate Distribution by Loan Outcome",
    subtitle = "Box plot with violin overlay | Outliers shown at 20% opacity",
    x = "Loan Outcome",
    y = "Initial Interest Rate (%)",
    fill = "Outcome"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "none"
  )
```

**Key Finding:** Charged-off loans carry measurably higher interest rates than loans paid in full. The median interest rate for defaulted loans sits roughly 1-2 percentage points above the paid-in-full median, and the interquartile range is wider — meaning defaulted loans show more variability in their pricing, not just higher rates on average.

Two interpretations are worth noting here. First, higher interest rates directly increase monthly debt service costs, reducing the borrower's cash flow buffer and increasing default probability — this is a direct causal mechanism. Second, lenders price risk into interest rates, meaning higher-rate loans were already flagged as riskier at origination. The interest rate therefore captures both the lender's ex-ante risk assessment and the borrower's ex-post repayment pressure simultaneously.

The violin shape for charged-off loans shows a second concentration of loans around 10-12%, likely corresponding to variable-rate loans that reset higher during the 2022-2023 rate hiking cycle identified in the previous chart. This connects the time series finding to the individual loan level — rising rates did not just affect aggregate default rates, they show up in the interest rate distribution of individual defaulted loans.

Interest rate will be included as a primary feature in Phase 4 modeling given this clear distributional separation between outcomes.

## Business Age vs Default Rate

```{r business-age-default}
#| label: business-age-default
#| fig-width: 10
#| fig-height: 6

# Calculate default rate by business age
business_age <- loans %>%
  filter(!is.na(business_age) & business_age != "NULL") %>%
  group_by(business_age) %>%
  summarise(
    total_loans = n(),
    total_defaults = sum(is_default),
    default_rate = round(mean(is_default) * 100, 2)
  ) %>%
  filter(total_loans > 100) %>%
  arrange(desc(default_rate))

# Clean up labels
business_age <- business_age %>%
  mutate(business_age = case_when(
    business_age == "NEW, LESS THAN 1 YEAR OLD"              ~ "New\n(<1 Year)",
    business_age == "NEW BUSINESS OR 2 YEARS OR LESS"        ~ "New\n(≤2 Years)",
    business_age == "STARTUP, LOAN FUNDS WILL OPEN BUSINESS" ~ "Startup\n(Pre-Revenue)",
    business_age == "EXISTING OR MORE THAN 2 YEARS OLD"      ~ "Existing\n(2+ Years)",
    business_age == "LESS THAN 3 YEARS OLD BUT AT LEAST 2"   ~ "Young\n(2-3 Years)",
    business_age == "LESS THAN 4 YEARS OLD BUT AT LEAST 3"   ~ "Young\n(3-4 Years)",
    business_age == "LESS THAN 5 YEARS OLD BUT AT LEAST 4"   ~ "Established\n(4-5 Years)",
    business_age == "EXISTING, 5 OR MORE YEARS"              ~ "Mature\n(5+ Years)",
    business_age == "CHANGE OF OWNERSHIP"                    ~ "Change of\nOwnership",
    business_age == "UNANSWERED"                             ~ "Unanswered",
    TRUE ~ business_age
)) %>%
  arrange(desc(default_rate)) %>%
  mutate(business_age = factor(business_age, levels = business_age))

# Plot
ggplot(business_age, aes(x = business_age,
                          y = default_rate,
                          fill = default_rate)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(default_rate, "%")),
            vjust = -0.5, size = 4) +
  scale_fill_gradient(low = "#FFF9C4", high = "#E65100") +
  scale_y_continuous(limits = c(0, max(business_age$default_rate) * 1.15)) +
  labs(
    title = "SBA Loan Default Rate by Business Age",
    subtitle = "Calculated from 489,935 loans | Groups with 100+ loans shown",
    x = "Business Age at Loan Origination",
    y = "Default Rate (%)",
    fill = "Default Rate"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "none"
  )
```

**Key Finding:** Business age shows a clear monotonic relationship with default risk — newer businesses default at significantly higher rates than established ones. New businesses under two years old default at 12.66%, more than double the 5.63% rate for mature businesses with five or more years of operating history. This nearly 2.4x multiplier makes business age one of the strongest categorical predictors in the dataset and confirms its inclusion as a primary feature in Phase 4 modeling.

Startups seeking pre-revenue financing show an 8.82% default rate — lower than businesses under one year old but higher than established firms. This is intuitive: startups that receive SBA financing have passed an initial underwriting screen, while businesses that have been operating for under a year have a demonstrated but short track record that provides limited predictive signal.

One data quality concern worth noting is the overlap between category definitions. The SBA data contains multiple labels that describe similar business ages — "New Business or 2 Years or Less", "New Less Than 1 Year Old", and "Existing or More Than 2 Years Old" create ambiguous boundaries in the 1-2 year range. This inconsistency reflects real-world data collection issues in federal reporting systems where category definitions changed across fiscal years. In Phase 3 feature engineering these overlapping categories will be consolidated into a single ordinal business age variable to prevent the model from treating semantically similar categories as independent signals.

## Correlation Matrix

```{r correlation-matrix}
#| label: correlation-matrix
#| fig-width: 10
#| fig-height: 8

# Select numeric columns for correlation analysis
numeric_cols <- loans %>%
  select(
    "Loan Amount"          = loan_amount,
    "SBA Guarantee %"      = sba_guarantee_pct,
    "Interest Rate"        = interest_rate,
    "Term (Months)"        = term_months,
    "Jobs Supported"       = jobs_supported,
    "Default"              = is_default
  ) %>%
  # Remove rows with any NA in these columns
  drop_na()

# Compute correlation matrix
cor_matrix <- cor(numeric_cols)

# Round for display
cor_matrix_rounded <- round(cor_matrix, 2)

# Plot
corrplot(
  cor_matrix,
  method = "color",
  type = "upper",
  order = "hclust",
  tl.col = "black",
  tl.srt = 45,
  tl.cex = 0.9,
  addCoef.col = "black",
  number.cex = 0.8,
  col = colorRampPalette(c("#B71C1C", "white", "#1565C0"))(200),
  title = "Correlation Matrix — SBA 7(a) Numeric Features",
  mar = c(0, 0, 2, 0)
)
```

**Key Finding:** Interest rate shows the strongest linear correlation with default at 0.13 — modest but meaningful given the binary nature of the target variable. Term months shows a negative correlation of -0.18 with default, suggesting longer term loans default less frequently — likely because longer terms reduce monthly payment pressure.

The most critical finding is the perfect 1.00 correlation between loan_amount and sba_guaranteed_amount. These two variables are mathematically redundant — the guaranteed amount is calculated directly from the loan amount using a fixed guarantee percentage. Keeping both in a model would cause severe multicollinearity, inflating standard errors in logistic regression and creating unstable coefficient estimates. sba_guaranteed_amount will be dropped entirely before Phase 4 modeling. sba_guarantee_pct is retained as it captures the guarantee structure independently of loan size.

Interest rate has a negative correlation of -0.26 with sba_guarantee_pct — loans with higher SBA guarantee percentages tend to have lower interest rates, which makes sense since the guarantee reduces lender risk and allows more favorable pricing. This interaction will be explored further in Phase 3 feature engineering.

All remaining numeric features show low enough inter-correlation to be retained as independent predictors in Phase 4.

## Summary of Findings

```{r summary-table}
#| label: summary-table

# Key findings summary table
summary_findings <- data.frame(
  Feature = c(
    "Business Age",
    "Interest Rate", 
    "Industry",
    "Geography",
    "Loan Amount",
    "Term Length",
    "Loan Volume Trend"
  ),
  Finding = c(
    "New businesses default at 12.66% vs 5.33% for mature — strongest categorical predictor",
    "Defaulted loans carry 1-2% higher rates — direct causal mechanism via debt service pressure",
    "Limousine services (18.75%) and residential remodeling (15.93%) lead default rates",
    "Florida leads at 11.74% — driven by hurricane exposure and high operating costs",
    "Smaller loans show higher relative default rates — but reflects underwriting selection bias",
    "Negative correlation with default (-0.18) — longer terms reduce monthly payment pressure",
    "2022-2023 default spike reflects Fed rate hikes — recent loans right-censored"
  ),
  Phase4_Action = c(
    "Consolidate overlapping categories into ordinal feature",
    "Include as primary numeric feature",
    "Use 2-digit NAICS sector as categorical feature",
    "Include state as categorical — flag for fairness analysis in Phase 5",
    "Use loan_size_bucket categorical — avoid raw amount due to skew",
    "Include as numeric feature",
    "Exclude 2023-2025 loans from training set or apply vintage weights"
  )
)

knitr::kable(
  summary_findings,
  col.names = c("Feature", "EDA Finding", "Phase 4 Action"),
  caption = "Summary of EDA Findings and Modeling Implications"
)
```

```{r disconnect}
#| label: disconnect
#| include: false

# Close database connection
dbDisconnect(conn)
```
