---
title: "SBA 7(a) Loan Data — Exploratory Analysis"
author: "Colin McBane"
date: today
format: html
execute:
  echo: true
  warning: false
  message: false
---

# SBA 7(a) Loan Data — Exploratory Analysis

## Overview

This report analyzes raw SBA 7(a) federal loan data cleaned in Phase 1. The goal is to understand the distribution of loan characteristics, identify default patterns, and inform feature selection for Phase 4 modeling.

```{r setup}
#| label: setup
#| message: false
#| warning: false

# Load libraries
library(tidyverse)
library(DBI)
library(RSQLite)
library(ggplot2)
library(corrplot)
library(scales)

# Connect to SQLite database built in Phase 1
conn <- dbConnect(
  RSQLite::SQLite(),
  "data/processed/sba_clean.db"
)

# Load the clean loans view into R
loans <- dbGetQuery(conn, "SELECT * FROM loans_clean")

# Quick sanity check
cat("Rows loaded:", nrow(loans), "\n")
cat("Columns:", ncol(loans), "\n")
cat("Default rate:", round(mean(loans$is_default) * 100, 2), "%\n")
```

## Loan Amount Distribution

```{r loan-distribution, message = False, warning = FALSE}
#| label: loan-distribution
#| fig-width: 10
#| fig-height: 6

# Loan amount distribution by default status
ggplot(loans, aes(x = loan_amount, fill = factor(is_default))) +
  geom_histogram(bins = 50, alpha = 0.7, position = "identity") +
  scale_x_continuous(
    labels = dollar_format(scale = 1e-6, suffix = "M"),
    limits = c(0, 2000000)
  ) +
  scale_fill_manual(
    values = c("0" = "#2196F3", "1" = "#F44336"),
    labels = c("0" = "Paid in Full", "1" = "Charged Off")
  ) +
  labs(
    title = "SBA 7(a) Loan Amount Distribution by Default Status",
    subtitle = "FY2010–Present | 489,935 loans",
    x = "Loan Amount",
    y = "Number of Loans",
    fill = "Outcome"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "top"
  )
```

**Key Finding:** SBA 7(a) loans are heavily concentrated below \$250,000 with a strong right skew. Charged-off loans appear proportionally more frequent in smaller loan buckets, which may suggest higher relative default risk at the lower end. However this is correlation not causation — larger loans undergo more rigorous underwriting scrutiny before approval, meaning the lower default rate on large loans likely reflects selection bias in the approval process rather than loan size itself being protective. This distinction will be important when interpreting model coefficients in Phase 4.

## Default Rate by Industry

```{r default-by-industry}
#| label: default-by-industry
#| fig-width: 12
#| fig-height: 7

# Pull industry summary from SQLite
industry <- dbGetQuery(conn, "
  SELECT naics_sector, industry_description, 
         total_loans, default_rate_pct
  FROM summary_by_industry
  WHERE total_loans > 500
  ORDER BY default_rate_pct DESC
  LIMIT 15
")

# Clean up industry descriptions for chart labels
# Clean up industry descriptions for chart labels
industry <- industry %>%
  group_by(industry_description) %>%
  slice_max(default_rate_pct, n = 1) %>%
  ungroup() %>%
  mutate(industry_description = str_wrap(industry_description, width = 40)) %>%
  distinct(industry_description, .keep_all = TRUE) %>%
  arrange(default_rate_pct) %>%
  mutate(industry_description = factor(industry_description,
                                        levels = unique(industry_description)))

# Plot
ggplot(industry, aes(x = industry_description, 
                      y = default_rate_pct,
                      fill = default_rate_pct)) +
  geom_col() +
  geom_text(aes(label = paste0(default_rate_pct, "%")),
            hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_fill_gradient(low = "#FFF9C4", high = "#B71C1C") +
  scale_y_continuous(limits = c(0, max(industry$default_rate_pct) * 1.15)) +
  labs(
    title = "Top 15 Industries by SBA Loan Default Rate",
    subtitle = "Industries with 500+ loans | Ranked by default rate",
    x = NULL,
    y = "Default Rate (%)",
    fill = "Default Rate"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "none"
  )
```

**Key Finding:** Limousine and personal transportation services show the highest default rates at nearly 19%, followed by residential remodeling and electronic retail. These industries share common risk characteristics — project-based or discretionary revenue, high sensitivity to economic cycles, and thin operating margins. This finding will inform industry-level risk features in Phase 4 modeling.

## Default Rate by State

```{r default-by-state}
#| label: default-by-state
#| fig-width: 12
#| fig-height: 7

# Pull state summary from SQLite
states <- dbGetQuery(conn, "
  SELECT borr_state, total_loans, default_rate_pct
  FROM summary_by_state
  WHERE total_loans > 200
  AND borr_state NOT IN ('', 'NULL')
  ORDER BY default_rate_pct DESC
  LIMIT 20
")

# Sort for chart
states <- states %>%
  arrange(default_rate_pct) %>%
  mutate(borr_state = factor(borr_state, levels = borr_state))

# Plot
ggplot(states, aes(x = borr_state,
                    y = default_rate_pct,
                    fill = default_rate_pct)) +
  geom_col() +
  geom_text(aes(label = paste0(default_rate_pct, "%")),
            hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_fill_gradient(low = "#E3F2FD", high = "#1565C0") +
  scale_y_continuous(limits = c(0, max(states$default_rate_pct) * 1.15)) +
  labs(
    title = "Top 20 States by SBA Loan Default Rate",
    subtitle = "States with 200+ loans | Ranked by default rate",
    x = NULL,
    y = "Default Rate (%)",
    fill = "Default Rate"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "none"
  )
```

**Key Finding:** Florida leads all states with an 11.74% default rate, 
followed by DC and New Jersey at roughly 10.6% and 10.6% respectively. 
Florida's elevated default rate is consistent with its exposure to 
natural disaster risk — hurricanes and flooding can devastate small 
businesses overnight, disrupting cash flow and making loan repayment 
impossible regardless of the borrower's original creditworthiness. 
This is a systemic geographic risk that individual underwriters cannot 
control at the loan level.

DC, New Jersey, and New York all appear in the top five, which aligns 
with high cost-of-living environments where small business operating 
costs — rent, labor, utilities — compress margins and reduce the buffer 
against economic shocks. A restaurant in Manhattan faces fundamentally 
different fixed cost pressure than the same restaurant in rural Tennessee.

Texas and California, despite their large loan volumes, show moderate 
default rates around 9-10%, suggesting that economic diversification 
at the state level provides some natural hedge against industry-specific 
downturns.

This geographic signal will be incorporated as a state-level risk feature 
in Phase 4 modeling. It also directly connects to Phase 5 fairness 
analysis — states with structurally higher default rates may create 
disparate impact concerns if geography is used as a direct model input.

## Default Rate Over Time

```{r default-by-year}
#| label: default-by-year
#| fig-width: 11
#| fig-height: 6

# Pull yearly summary from SQLite
yearly <- dbGetQuery(conn, "
  SELECT approval_year, total_loans, 
         default_rate_pct, avg_loan_amount
  FROM summary_by_year
  WHERE approval_year IS NOT NULL
  ORDER BY approval_year
")

# Dual axis plot — default rate and loan volume over time
ggplot(yearly, aes(x = approval_year)) +
  geom_col(aes(y = total_loans / 10),
           fill = "#E3F2FD", color = "#90CAF9", alpha = 0.7) +
  geom_line(aes(y = default_rate_pct * 100),
            color = "#B71C1C", linewidth = 1.2) +
  geom_point(aes(y = default_rate_pct * 100),
             color = "#B71C1C", size = 3) +
  scale_x_continuous(breaks = seq(min(yearly$approval_year),
                                   max(yearly$approval_year), 1)) +
  scale_y_continuous(
    name = "Loan Volume (scaled)",
    sec.axis = sec_axis(~ . / 100,
                        name = "Default Rate (%)",
                        labels = percent_format())
  ) +
  labs(
    title = "SBA 7(a) Loan Volume and Default Rate by Approval Year",
    subtitle = "Blue bars = loan volume | Red line = default rate",
    x = "Approval Year"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

**Key Finding:** This chart reveals three distinct macroeconomic periods 
that shaped SBA 7(a) lending patterns between 2010 and 2026.

**2010–2019 — Post-Crisis Recovery:** Loan volume grew steadily through 
the decade as the economy recovered from the 2008 financial crisis. 
Default rates remained relatively stable in the 1,000–1,500 basis point 
range, reflecting a healthy but cautious lending environment.

**2020–2021 — COVID Displacement:** Loan volume dropped sharply in 2020 
as the SBA redirected resources toward emergency programs — the Paycheck 
Protection Program (PPP) and Economic Injury Disaster Loans (EIDL) — which 
provided direct grants and forgivable loans to small businesses. These 
programs temporarily absorbed demand that would otherwise have flowed 
through the 7(a) program. Critically, default rates did not spike 
immediately during the pandemic because government relief programs 
artificially supported borrower cash flows, masking underlying credit stress.

**2022–2023 — The Macroeconomic Double Whammy:** Default rates peaked in 
this period as the Federal Reserve aggressively raised interest rates to 
combat post-pandemic inflation — the fastest rate hiking cycle since the 
1980s. Borrowers who took on variable-rate SBA loans during the low-rate 
environment of 2020–2021 suddenly faced materially higher debt service 
costs. Simultaneously, inflation compressed operating margins as input 
costs rose faster than businesses could pass through price increases to 
customers. The combination of higher rates and margin compression created 
a perfect storm for small business defaults — explaining why the default 
spike lagged the pandemic by two years rather than occurring during it.

**2024–2026 — Right Censoring:** Default rates appear to fall sharply in 
recent years, but this is a statistical artifact rather than a genuine 
improvement in credit quality. Loans approved in 2023–2025 have not had 
sufficient time to season and default. In credit risk modeling this is 
called right censoring — recent observations are incomplete because the 
outcome has not yet been observed. Phase 4 modeling will account for this 
by weighting or excluding recent vintages from the training set to avoid 
biasing the model toward artificially low default rates in recent years.
